---
title: "Analysis of American Music Industry Sales from 1975-2019"
author: "Patton Nelson"
date: "2023-12-19"
output:
  word_document:
    toc: yes
  html_document:
    theme: journal
    toc: yes
    toc_float:
      collapsed: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,include = FALSE}
library(gapminder)
library(rmdformats)
library(here)
library(socviz)
library(ggplot2)
library(tidyverse)
library(forcats)
library(dplyr)
library(scales)
library(data.table)
library(ggsankey)
library(plotly)
library(knitr)
library(showtext)
library(extrafont)
```

```{r, include = FALSE}
#initial construction and manipulation of all dataframes needed for the following plots.
music <- readr::read_csv('MusicData.csv') %>%
  mutate(format = recode(format, 'Cassette Single'='Cassette',
                         'CD Single'='CD',
                         'Download Album'='Download',
                         'Download Music Video'='Download',
                         'Download Single'='Download',
                         'Limited Tier Paid Subscription' = 'Paid Subscription')) %>%
  filter(as.numeric(year) >= 1975)
units <- subset(music, metric == 'Units') %>%
  aggregate(value_actual ~ format + year, sum) %>%
  arrange(format,year) %>%
  rename(Format = format,
         Year = year,
         Units = value_actual)
revenue <- subset(music, metric == 'Value') %>%
  aggregate(value_actual ~ format + year, sum) %>%
  arrange(format,year) %>%
  filter(value_actual > 100) %>%
  rename(Format = format,
         Year = year,
         Revenue = value_actual)
revenue_adjusted <- subset(music, metric == 'Value (Adjusted)') %>%
  aggregate(value_actual ~ format + year, sum) %>%
  arrange(format,year) %>%
  rename(Format = format,
         Year = year,
         Revenue = value_actual)
```

```{r, include = FALSE}
#Creating Units Plot
Format <- units$Format
unitsplot <- ggplot(units, mapping = aes(x=Year, y = Units)) + 
  geom_path(aes(color = Format), size = 0.5) + 
  geom_area(position = position_identity(), alpha = .20, aes(fill = Format), stat = 'identity') +
  scale_x_discrete(breaks = seq(1975,2020,5),
                   limits = seq(1975,2020,5),
                   expand = c(0,0)) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white",
                                  colour = "black",
                                  size = 0.5,
                                  linetype = "dotted"),
    panel.grid.major = element_line(color = 'lightgrey')) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  xlab('Year') +
  ylab('Units')
```

```{r, include = FALSE}
#Creating dataframe needed for sankey chart
sankey <- revenue_adjusted %>%
  aggregate(Revenue ~ Format, sum) %>%
  mutate(Form = case_when(Format == '8 - Track'  ~ "Physical",
                          Format == 'Cassette' ~ 'Physical',
                          Format == 'CD' ~ 'Physical',
                          Format == 'DVD Audio' ~ 'Physical',
                          Format == 'Kiosk' ~ 'Physical',
                          Format == 'LP/EP' ~ 'Physical',
                          Format == 'Music Video (Physical)' ~ 'Physical',
                          Format == 'Other Tapes' ~ 'Physical',
                          Format == 'SACD' ~ 'Physical',
                          Format == 'Vinyl Single' ~ 'Physical',
                          Format == 'Download' ~ 'Digital',
                          Format == 'On-Demand Streaming (Ad-Supported)' ~ 'Digital',
                          Format == 'Other Ad-Supported Streaming' ~ 'Digital',
                          Format == 'Other Digital' ~ 'Digital',
                          Format == 'Paid Subscription' ~ 'Digital',
                          Format == 'Ringtones & Ringbacks' ~ 'Digital',
                          Format == 'SoundExchange Distributions' ~ 'Digital',
                          Format == 'Synchronization' ~ 'Digital')) %>%
  arrange(Revenue) %>%
  add_column(Total = 'Revenue') %>%
  make_long(Total, Form, Format) |> 
  mutate(node = forcats::fct_inorder(node))
```

```{r, include = FALSE}
#Begin building sankey chart
sank <- ggplot(sankey, aes(x = x
                     , next_x = next_x
                     , node = node
                     , next_node = next_node
                     , fill = node
                     , label = node)
             )
sank <- sank +geom_sankey(flow.alpha = 0.90
                      , node.color = "black"
                      ,show.legend = FALSE) +
  geom_sankey_label(size = 3, color = "black", fill= "white", hjust = 0) + theme_bw() +
  theme(legend.position = "none") + 
  theme_sankey(base_size = 12) +
  theme(axis.title = element_blank(),
        axis.text.y = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())+
  labs(caption = '*Formats are ranked top-to-bottom from highest-to-lowest contribution \n to overall revenue (inflation adjusted) from years 1975 - 2020') +
  labs(fill = 'Nodes') +
  scale_fill_manual(values = c('Digital' = '#339966',
                               'Physical' = '#336699',
                               'Revenue' = '#330066',
                               'CD' = '#006699',
                               'Cassette' = '#006699',
                               'LP/EP' = '#006699',
                               '8 - Track' = '#006699',
                               'Vinyl Single' = '#006699',
                               'Music Video (Physical)' = '#006699',
                               'SACD' = '#006699',
                               'DVD Audio' = '#006699',
                               'Other Tapes' = '#006699',
                               'Kiosk' = '#006699',
                               'Download' = '#33cc33',
                               'Paid Subscription' = '#33cc33',
                               'SoundExchange Distributions' = '#33cc33',
                               'Ringtones & Ringbacks' = '#33cc33',
                               'On-Demand Streaming (Ad-Supported)' = '#33cc33',
                               'Synchronization' = '#33cc33',
                               'Other Ad-Supported Streaming' = '#33cc33',
                               'Other Digital' = '#33cc33'))
  


```

## Format Breakdown - Physical & Digital
```{r, echo = FALSE, warning = FALSE}
sank
```

## Unit Sales
```{r, echo = FALSE}
ggplotly(unitsplot)
```

Note : Units are defined by the RIAA (Recording Industry Association of America) as to equal 1 album sale, 15 track sales, and 1500 streams


## Revenue {.tabset}
### Not Adjusted
```{r, include = FALSE}
#Initial Revenue Plot. Not adjusted for inflation
revplot <- ggplot(revenue, mapping = aes(x=Year, y = Revenue)) + 
  geom_area(position = position_identity(), alpha = .35, aes(fill = revenue$Format), stat = 'identity') +
  geom_path(aes(color = Format)) + 
  scale_x_discrete(breaks = seq(1975,2020,5),
                   limits = seq(1975,2020,5),
                   expand = c(0,0)) +
  scale_color_viridis_d() + 
  scale_fill_viridis_d() +
  theme_bw() + 
  theme(panel.grid = element_line(color = 'grey',
                                  linetype = 'dotted',
                                  size = 0.5)) +
  guides(fill = FALSE) +
labs(title = 'Revenue (in millions) accrued by Format from 1975 - 2019')
```

```{r, echo = FALSE, warning=FALSE}
ggplotly(revplot)
```

### Adjusted for inflation
```{r, include = FALSE}
#Making a second plot to be included for adjusted revenue.
adjustedplot <- ggplot(revenue_adjusted, mapping = aes(x=Year, y = Revenue)) + 
  geom_area(position = position_identity(), alpha = .35,aes(fill = revenue_adjusted$Format), stat = 'identity') +
  geom_path(aes(color = Format)) + 
  scale_x_discrete(breaks = seq(1975,2020,5),
                   limits = seq(1975,2020,5),
                   expand = c(0,0)) +
  theme(
    panel.background = element_rect(fill = "white",
                                  colour = "black",
                                  size = 1,
                                  linetype = "solid"),
    panel.grid = element_line(color = 'grey',
                              size = 0.5,
                              linetype = 'dotted')) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  guides(fill = FALSE) + 
  labs(title = 'Revenue (adjusted for inflation) Accrued by Format from 1975 - 2019')
```
```{r, echo = FALSE, warning = FALSE}
#Revenue Adjusted Plot
ggplotly(adjustedplot)
```

```{r, include = FALSE}
#Creating dataframe from revenue adjusted for a pie chart. I used revenue adjusted as I wanted to get an accurate comparison of the revenue in total from 1975-2019
pie_df <- revenue_adjusted %>%
  aggregate(Revenue ~ Format, sum)
others <- subset(pie_df, Format %in% c('DVD Audio', 'Kiosk','Other Digital','Other Tapes','SACD', 'Synchronization','SoundExchange Distributions','Ringtones & Ringbacks', 'Other Ad-Supported Streaming','On-Demand Streaming (Ad-Supported)')) %>%
  arrange(Revenue)
pie_df <- pie_df %>%
  mutate(Format = recode(Format,'DVD Audio'='Other',
                         'Kiosk' = 'Other',
                         'Other Digital' = 'Other',
                         'Other Tapes' = 'Other',
                         'SACD' = 'Other',
                         'Synchronization' = 'Other',
                         'SoundExchange Distributions' = 'Other',
                         'Ringtones & Ringbacks' = 'Other',
                         'Other Ad-Supported Streaming' = 'Other',
                         'On-Demand Streaming (Ad-Supported)' = 'Other')) %>%
  aggregate(Revenue ~Format, sum)
#PIE CHART
pie <- ggplot(pie_df, aes(x="", y=Revenue, fill=Format)) +
  geom_bar(stat="identity", width=1, color = 'white') +
  coord_polar("y", start=0) +
  scale_fill_viridis_d(aesthetics = 'fill') +
  theme_void() +
  guides(fill=guide_legend(title="Format")) 
```
## Revenue Breakdown by Format

```{r, echo = FALSE, warning = FALSE}
pie
```


-Assessed from total revenue (adjusted for inflation) years 1975-2019

- "Other" section comprises of DVD Audio, Kiosk, Other Digital, Other Tapes, SACD, Synchronization, SoundExchange Distributions, Ringtones & Ringbacks, Other Ad-Supported Streaming, On-Demand Streaming (Ad-Supported).

## Comments
With comparison of revenue vs unit sales, we can observe overall trends in the lifespan of the American Music Industry. The industry first boomed in the early 1970's with physical LPs being the leading format to bring in revenue. Since then we see the rise and fall of various formats of music. A large peak in both revenue and unit sales signifies the height of the American Music Industry in the late 1990's - early 2000's. Unit sales reached an all-time peak in 2012 from Downloads, but the contrast between the unit-to-revenue ratio in 1999 to 2012 highlights the decrease in monetary value for units over the past 20 years. Additionally, the rapid increase in unit sales from 2015 - 2019 from Paid Subscriptions in contrast its revenue displays the same decrease in monetary value for unit sales.

*Using public data from RIAA
